Concurrent Programming in Harmony

### 2021-04-18 book/paper.tex
New chapter: debugging

\begin{itemize}
\item \autoref{ch:harmonyintro} introduces the Harmony programming
language, as it provides the language for presenting synchronization
problems and solutions.
\item \autoref{ch:concurrent} illustrates the problem of
concurrent programming through a simple example in which two threads
are concurrently incrementing a counter.
\item \autoref{ch:harmonymachine} presents the
Harmony virtual machine to understand the problem
underlying concurrency better.
\item \autoref{ch:critical} introduces the concept of a
\emph{critical section} and presents various flawed implementations
of critical sections to demonstrate that implementing a critical section
is not trivial.
\item \autoref{ch:peterson} introduces \emph{Peterson's Algorithm}, an
elegant (although not very efficient or practical) solution to implementating a critical section.
\item \autoref{ch:method} gives some more details on the Harmony
language needed for the rest of the book.
\item \autoref{ch:spinlock} introduces atomic \emph{locks}
for implemented critical sections.
\item \autoref{ch:synch} presents Harmony modules
that support locks and other synchronization primitives.
\item \autoref{ch:cds} gives an introduction to building concurrent
data structures.
\item \autoref{ch:testing} discusses approaches to testing
concurrent code in Harmony.
\item \autoref{ch:condwait} talks about threads having to wait for
certain conditions.  As examples, it presents the reader/writer lock
problem and the bounded buffer problem.
\item \autoref{ch:sbs} presents \emph{Split Binary Semaphores}, a
general technique for solving synchronization problems.
\begin{comment}
\autoref{ch:rdwrbusy} also discusses \emph{busy-waiting}, which is
an easy but undesirable approach to synchronize threads.
\item \autoref{ch:semaphore} introduces \emph{semaphores},
a generalization of locks
that is good not only for mutual exclusion but also for waiting for
certain application-level conditions.
\item \autoref{ch:bb} and \autoref{ch:sbs} introduce the
\emph{bounded buffer problem} (aka \emph{producer/consumer problem})
and various solutions.
\end{comment}
\item \autoref{ch:starvation} talks about \emph{starvation}:
the problem that in some
synchronization approaches threads may not be able to get access to a
resource they need.
\item \autoref{ch:monitors} presents
\emph{monitors} and \emph{condition variables},
another approach to thread synchronication.
\item \autoref{ch:deadlock} describes \emph{deadlock}
where a set of threads are indefinitely waiting for one another to
release a resource.
\item \autoref{ch:actor} presents the \emph{actor model}
and \emph{message passing} as an approach to synchronization.
\item \autoref{ch:barrier} describes \emph{barrier synchronization},
useful in high-performance computing applications such as parallel simulations.
\item \autoref{ch:interrupts} discusses how to handle interrupts,
a problem closely related to---but not the same as---synchronizing threads.
\item \autoref{ch:nonblocking} introduces \emph{non-blocking} or
\emph{wait-free} synchronization algorithms,
which prevent threads waiting for one another more than a bounded number of
steps.
\item \autoref{ch:abp} presents a problem and a solution to the distributed
systems problem of having two threads communicate reliably over an unreliable
network.
\end{itemize}

@@

>>>>

\begin{itemize}
\item \autoref{ch:harmonyintro} introduces the Harmony programming
language, as it provides the language for presenting synchronization
problems and solutions.
\item \autoref{ch:concurrent} illustrates the problem of
concurrent programming through a simple example in which two threads
are concurrently incrementing a counter.
\item \autoref{ch:harmonymachine} presents the
Harmony virtual machine to understand the problem
underlying concurrency better.
\item \autoref{ch:critical} introduces the concept of a
\emph{critical section} and presents various flawed implementations
of critical sections to demonstrate that implementing a critical section
is not trivial.
\item \autoref{ch:peterson} introduces \emph{Peterson's Algorithm}, an
elegant (although not very efficient or practical) solution to implementating a critical section.
\item \autoref{ch:method} gives some more details on the Harmony
language needed for the rest of the book.
\item \autoref{ch:spinlock} introduces atomic \emph{locks}
for implemented critical sections.
\item \autoref{ch:synch} presents Harmony modules
that support locks and other synchronization primitives.
\item \autoref{ch:cds} gives an introduction to building concurrent
data structures.
\item \autoref{ch:testing} discusses approaches to testing
concurrent code in Harmony.
\item \autoref{ch:debugging} instead goes into how to find a bug
in concurrent code using the Harmony output.
\item \autoref{ch:condwait} talks about threads having to wait for
certain conditions.  As examples, it presents the reader/writer lock
problem and the bounded buffer problem.
\item \autoref{ch:sbs} presents \emph{Split Binary Semaphores}, a
general technique for solving synchronization problems.
\begin{comment}
\autoref{ch:rdwrbusy} also discusses \emph{busy-waiting}, which is
an easy but undesirable approach to synchronize threads.
\item \autoref{ch:semaphore} introduces \emph{semaphores},
a generalization of locks
that is good not only for mutual exclusion but also for waiting for
certain application-level conditions.
\item \autoref{ch:bb} and \autoref{ch:sbs} introduce the
\emph{bounded buffer problem} (aka \emph{producer/consumer problem})
and various solutions.
\end{comment}
\item \autoref{ch:starvation} talks about \emph{starvation}:
the problem that in some
synchronization approaches threads may not be able to get access to a
resource they need.
\item \autoref{ch:monitors} presents
\emph{monitors} and \emph{condition variables},
another approach to thread synchronication.
\item \autoref{ch:deadlock} describes \emph{deadlock}
where a set of threads are indefinitely waiting for one another to
release a resource.
\item \autoref{ch:actor} presents the \emph{actor model}
and \emph{message passing} as an approach to synchronization.
\item \autoref{ch:barrier} describes \emph{barrier synchronization},
useful in high-performance computing applications such as parallel simulations.
\item \autoref{ch:interrupts} discusses how to handle interrupts,
a problem closely related to---but not the same as---synchronizing threads.
\item \autoref{ch:nonblocking} introduces \emph{non-blocking} or
\emph{wait-free} synchronization algorithms,
which prevent threads waiting for one another more than a bounded number of
steps.
\item \autoref{ch:abp} presents a problem and a solution to the distributed
systems problem of having two threads communicate reliably over an unreliable
network.
\end{itemize}

@@

\chapter{Debugging}
\label{ch:debugging}

\begin{figure}
\begin{code}
\harmonysource{queuebug}
\end{code}
\caption{\harmonylink{code/queuebug.hny} A buggy queue implementation.}
\label{fig:queuebug}
\end{figure}

\begin{figure}
\begin{center}
\includegraphics[width=\textwidth]{figures/queuefix1.pdf}
\end{center}
\caption{Harmony output.}
\label{fig:queuefix1}
\end{figure}

So you wrote a Harmony program and Harmony reports a bug.
Often you may just be able to figure it out by staring at the code and
going through some easy scenarios, but what if you don't?
The output of Harmony can be helpful in that case.

\autoref{fig:queuebug} contains an attempt at a queue implementation
where the queue is implemented by a linked list, with the first node
being a \texttt{dummy} node to prevent data races.  Each node in the
list contains a lock.  The \texttt{put}() method walks the list until
it gets to the last node, each time acquiring the lock to access the
node's fields.  When \texttt{put}() gets to the last node in the list,
it appends a new one.  The \texttt{get}() method locks the first (dummy)
node, removes the second from the list and frees it.  The method returns
the value from the removed node.

Let us run the code through some of the test programs in the last
chapter.  Harmony does not detect any issues with the sequential test
in \autoref{fig:qtestseq}.  The concurrent \texttt{put} test of
\autoref{fig:qtest1} as well as the concurrent \texttt{get} test of
\autoref{fig:qtest2} also find no problems.  However, when we run
the new queue code through the test in \autoref{fig:qtest3}, Harmony
finds a problem.

\autoref{fig:queuefix1} shows the Harmony output of running
the test in \autoref{fig:qtest3} against the queue code
in \autoref{fig:queuebug}.
There is quite a bit of information in the Harmony output, and it's
important to learn to navigate through it.
Let's start with looking at the red text.  Harmony found a safety
violation (something bad happened during one of the possible executions),
and in particular \texttt{putter}(1) (thread T2) was trying to dereference
\texttt{alloc}.\textit{pool}[0].\texttt{lock} in turn 5.

The \texttt{alloc} module maintains a shared array \textit{pool} that
it uses for dynamic allocation.  Apparently \texttt{putter}(1) tried to access
\textit{pool}[0], but it does not exist, meaning that either it was not
yet allocated, or it had been freed since it was allocated.
When we look at the top half of the figure, we see that in fact
\texttt{putter}(0) (T1) allocated \textit{pool}[0] in turn 2, while
\texttt{getter}() (T4) released it in turn 4.

So how did we get there?  We can start by single stepping through the
actions of \texttt{putter}(0) by clicking on its first block.
By hitting return repeatedly, we can go through the lines of code that
it is executing.  Doing so, we can see that it allocates
\textit{pool}[0] and uses that as the node to contain value 0 and add
this node to the queue \textit{qtest}.  At the end of its turn,
\texttt{putter}(0) has just finished \texttt{put}(0).  Looking at the
state in the top right, everything looks good.  The first node
(\textit{testq}) points to the allocated node with the value 0 in it,
and its \texttt{next} pointer is \textbf{None}.  The locks on both nodes
are released.  So far so good.

Next \texttt{putter}(1) (thread T2) takes a turn.  It gets to line 10 in
\autoref{fig:queuebug} where it's trying to acquire
$q$$\rightarrow$\texttt{lock}.  If we look at the bottom right,
we see that $q$ points to \textit{pool}[0], the node that has value 0
in it.  Again, so far so good.

But before \texttt{putter}(1) obtains
the lock on \textit{pool}[0], \texttt{getter}() (thread T4) starts
running.  \texttt{getter}() acquires the lock on \textit{qtest}, which
still points to \textit{pool}[0].  \texttt{getter} then extracts the
value from \textit{pool}[0] and then releases it.
At this point \textit{pool}[0] no longer exists, but then
\texttt{putter}(1) starts running again from where it left off.
It was about to acquire the lock on \textit{pool}[0], but now that
node no longer exists.

To fix the code, we have to use hand-over-hand locking (\autoref{ch:cds}).
\autoref{fig:queuefix} shows an implementation that uses the same
data structure as \autoref{fig:queuebug} but uses hand-over-hand locking,
both for \texttt{put}() and for \texttt{get}().  It passes all tests.

\begin{figure}
\begin{code}
\harmonysource{queuefix}
\end{code}
\caption{\harmonylink{code/queuefix.hny} Queue implementation with hand-over-hand locking.}
\label{fig:queuefix}
\end{figure}
